Making Sequential Programs Deterministic

### 2019-04-05 randoop-case-study.tex
Expanding case study

%% This is not relevant in this section of the paper, where no times are given.
% We ran all our
% experiments on a 2.3 GHz Intel Core i5 processor with 8 GB of RAM.

As a case study, we applied \theDeterminismChecker to the Randoop test
generation tool~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We chose Randoop because 
it is frequently used in software engineering experiments,
it is actively maintained,
and its developers have struggled with nondeterminism (personal
communication, 2019).

Randoop is intended to be deterministic, when invoked on a deterministic
program.\footnote{Users of Randoop can pass in a different seed in order to
  obtain a different deterministic output.  Randoop has command-line
  options that enable concurrency and timeouts, both of which can lead to
  nondeterministic behavior.}

However, Randoop was not deterministic.  This caused the developers
problems in 
reproducing bugs reported by users, 
reproducing test failures during development, and
understanding the effect of changes to Randoop (by comparing executions of two
similar variants of Randoop).

The developers took extensive action to detect and mitigate the effect of nondeterminism.
They used Docker images to run tests, to avoid system dependencies such as
a different JDK having a different number of classes or methods.
They wrote tests with relaxed oracles (assertions) that permit multiple
possible answers --- for example, in code coverage of generated tests.
They used linters such as Error Prone, which warns if \<toString> is used on
array objects, which don't override \<Object.toString> and therefore print a
hash code which may vary from run to run.
They used libraries that make hash codes deterministic, by giving each
object of a type a unique ID that counts up from 1, rather than using a
memory address as \<Object.hashCode> does.
They wrote preprocess output and logs to make nondeterministic ones easier
to compare, such as by removing or canonicalizing hash codes, dates, and
other nondeterministing output.
Each of these efforts was helpful, but they were not enough to solve the problem.

In July 2017, the Randoop developers spent two weeks of full-time work to
eliminate unintentional nondeterministic behavior in Randoop.
Their methodology was to repeatedly
 run Randoop with verbose logging enabled,
 look for differences in logging output,
 find the root cause of nondeterminism,
 and eliminate it.
The most common causes were \<toString()> routines and iteration order of sets and maps.
Some of the nondeterminism was in libraries, such as the JDK\@.

The developers chose not to make every \<Map> a \<LinkedHashMap>, because that
would have increased memory and CPU costs.  They chose not to make every
order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
not always needed and would have increased costs.

\todo{When we discuss annotation effort, compare to this.
  The annotation effort is
  high, but it's no more than the hundreds of hours that the Randoop
  developers spent, with less success.}

That coding sprint did not find all the problems.
The developers debugged and fixed 5 additional determinism defects over the
next 12 months, using a similar methodology.
% [2413 changed lines in the diffs.
% (This number slightly overstates the number of code changes made, because sometimes changing code would result in code formatting changes, and some changes in those commits are unrelated to determinism, such as documentation improvements.)]
We analyzed a version of Randoop from February 2019 (commit 0a8e63fb), over 6 months after the most recent determinism bug fix.

@@

\TheDeterminismChecker found three previously-unknown nondeterminism bugs in Randoop.
The Randoop developers accepted our bug reports and committed fixes to the repository. A summary
of these bugs follows:

\begin{itemize}
\item \bugOne:
  A use of \<HashSet> that the developers changed to \<LinkedHashSet>\commit{c975a9f7}.
The old code could cause a problem if a type variable's lower or upper
bound has a type parameter that the type variable itself does not have.
This situation does occur, even in Randoop's test suite.

\item \bugTwo:
Dependency on the CLASSPATH environment variable in preference to the
classpath passed on the command line\commit{330e3c56}.
This could have led to incorrect behavior if a user set the environment variable.
The developers changed Randoop to not read the environment variable.

\item \bugThree:
Diagnostic output that printed a hash code for brevity, but should have had
deterministic output\commit{661a4970}.  By contrast to the first two bugs,
this one is not serious.

\item \bugFour:
Diagnostic output that printed a timestamp, but should have obeyed an
option about whether to do so\commit{a460df97}.  By contrast to the first
two bugs, this one is not serious.

\end{itemize}

@@

\begin{Verbatim}
55:-    String classPath = "";
55:+    String classPath = JUNIT_JAR;
56:     if (dependencies != null) {
57:       for (String s : dependencies) {
58:         Path file = Paths.get(s);
59:         classPath += (pathSeparator + file.toAbsolutePath().toString());
60:       }
61:     }
\end{Verbatim}

\caption{Fixes made by the Randoop developers in response to our bug report
  about use of environment variables.  Lines starting with ``\<->'' were
  removed and those starting with ``\<+>'' were added.
  \TheDeterminismChecker verified all other uses of system and Java properties.}
\label{fig:randoop-bug-getenv}
\end{figure}

@@

@@

>>>>

To evaluate the usability of \theDeterminismChecker,
we applied it to the Randoop test
generation tool~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We also wrote specifications for libraries it uses, such as the JDK, JUnit,
and others.
Like \theDeterminismCheckerImplementation,
all the case study materials are publicly available at \myurl{\todo{url for
    case study}} for reproducibility.

\subsection{Subject program}

We chose Randoop because 
it is frequently used in software engineering experiments,
it is actively maintained,
and its developers have struggled with nondeterminism~\cite{randoop-issue-tracker,randoop-mailing-lists}.

Randoop is intended to be deterministic, when invoked on a deterministic
program~\cite{randoop-manual}.\footnote{Users of Randoop can pass in a different seed in order to
  obtain a different deterministic output.  Randoop has command-line
  options that enable concurrency and timeouts, both of which can lead to
  nondeterministic behavior.}
However, Randoop was not deterministic.  This caused the developers
problems in 
reproducing bugs reported by users, 
reproducing test failures during development, and
understanding the effect of changes to Randoop by comparing executions of two
similar variants of Randoop.

The developers took extensive action to detect and mitigate the effect of nondeterminism.
They used Docker images to run tests, to avoid system dependencies such as
a different JDK having a different number of classes or methods.
They wrote tests with relaxed oracles (assertions) that permit multiple
possible answers --- for example, in code coverage of generated tests.
They used linters such as Error Prone, which warns if \<toString> is used on
array objects, which don't override \<Object.toString> and therefore print a
hash code which may vary from run to run.
They used a library that makes hash codes deterministic, by giving each
object of a type a unique ID that counts up from 1 rather than using a
memory address as \<Object.hashCode> does.
They wrote specialized tools to preprocess output and logs to make them easier
to compare, such as by removing or canonicalizing hash codes, dates, and
other nondeterministing output (commits e15f9155--b473fd14).
Each of these efforts was helpful, but they were not enough to solve the problem.

In July 2017, the Randoop developers spent two weeks of full-time work to
eliminate unintentional nondeterministic behavior in Randoop.
Their methodology was to repeatedly
 run Randoop with verbose logging enabled,
 look for differences in logging output,
 find the root cause of nondeterminism,
 and eliminate it.
Some of the nondeterminism was in libraries, such as the JDK\@.
The most common causes were \<toString> routines and iteration order of sets and maps.
The most common fixes were to change the implementation of \<toString>
and to use \<LinkedHashSet> and \<LinkedHashMap> or to sort collections
before iterating over them.
The developers did not make every \<Set> and \<Map> a \<LinkedHashSet> or \<LinkedHashMap>, because that
would have increased memory and CPU costs.  They chose not to make every
order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
not always needed and would have increased costs.

That coding sprint did not find all the problems.
The developers debugged and fixed 5 additional determinism defects over the
next 12 months, using a similar methodology
(commits c15ccbf2, 44bdeebd, 5ff5b4c4, 22eda87f, and b473fd14).

% [2413 changed lines in the diffs.
% (This number slightly overstates the number of code changes made, because sometimes changing code would result in code formatting changes, and some changes in those commits are unrelated to determinism, such as documentation improvements.)]
We analyzed a version of Randoop from late February 2019 (commit 0a8e63fb), over 6 months after the most recent determinism bug fix.

\subsection{Methodology}

In brief, we wrote type qualifiers in the Randoop source code to express its
determinism specification
We then ran
\theDeterminismChecker.  Each warning indicated a mismatch between the
specification and the implementation, which we addressed by changing our
specification or reporting a bug in the implementation.

@@

\subsection{Results}

\TheDeterminismChecker found \numRandoopBugs previously-unknown nondeterminism bugs in Randoop.
The Randoop developers accepted our bug reports and committed fixes to the repository. A summary
of these bugs follows, according to the Randoop developers' categorization:

\begin{description}
\item[Severe issues]
  \strut
\begin{description}
\item[\bugHashSet]
One use of \<HashSet> could cause a problem if a type variable's lower or upper
bound has a type parameter that the type variable itself does not have.
This situation does occur, even in Randoop's test suite.
The developers fixed this by changing a \<HashSet> to \<LinkedHashSet>
(commit c975a9f7, shown in \cref{fig:randoop-bug-hashset}).
\TheDeterminismChecker confirmed that 
25 other uses of \<new HashSet> were acceptable, as were 15 uses of \<new HashMap>.

\item[\bugClasspath]
Randoop used the CLASSPATH environment variable in preference to the
classpath passed on the command line.
This could have led to incorrect behavior, both in Randoop's test suite and in the field,
if a user set the environment variable.
The developers fixed it by changing Randoop to not read the environment variable
(commit 330e3c56, shown in \cref{fig:randoop-bug-getenv}).
\TheDeterminismChecker verified that all other uses of system and Java
properties did not lead to nondeterministic behavior.
\end{description}

\item[Minor issues]
\strut
\begin{description}

\item \bugHashcodeOutput:
  Diagnostic output printed a hash code for brevity.
  The developers changed it to have deterministic output (commit 661a4970).
  Although the Randoop developers classified this issue as minor, it is
  very similar to ones they fixed during their two-week coding sprint.

\item \bugTimestampOutput:
  Diagnostic output printed a timestamp.
  The developers fixed it by making it obeyed an
  option about whether to print timestamps (commit a460df97).
  The developers may not have noticed this because their log-postprocessing
  tools removed them from the log when doing comparisons.

\end{description}

\end{description}

@@

\begin{Verbatim}
55:-    String classPath = "";
55:+    String classPath = JUNIT_JAR;
56:     if (dependencies != null) {
57:       for (String s : dependencies) {
58:         Path file = Paths.get(s);
59:         classPath += (pathSeparator + file.toAbsolutePath().toString());
60:       }
61:     }
\end{Verbatim}
\caption{Fixes made by the Randoop developers in response to our bug report
  about use of environment variables.  Lines starting with ``\<->'' were
  removed and those starting with ``\<+>'' were added.
  \TheDeterminismChecker verified all other uses of system and Java properties.}
\label{fig:randoop-bug-getenv}
\end{figure}

@@

\subsection{False positive warnings}

Java 8 lambdas.

\subsection{Annotation effort}

@@

\todo{When we discuss annotation effort, compare to the manual work.
  The annotation effort is
  high, but it's no more than the hundreds of hours that the Randoop
  developers spent, with less success.}
